SIR-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-03-13 paper_draft/main.tex
small fixes

The need for a nuanced understanding of how interventions impact both health outcomes and economic indicators became increasingly evident, prompting a comprehensive examination by epidemiologists to assist policy makers~\cite{Anderson2020}. The outbreak of COVID-19 has prompted epidemiologists to research on various aspects, including mobility control~\cite{Song2022,Chinazzi2020}, vaccination strategies~\cite{Nguyen2021, Kim2022}, non-pharmaceutical interventions (NPIs) like restricting population movements and gatherings, closing schools and businesses, and requiring masks indoors~\cite{Jalloh2022,Caldwell2021,Ferguson2020}, and financial considerations~\cite{DeFoo2023}. Despite the numerous studies conducted, very few explore how common interventions meet multiple policy objectives or how a precise articulation of the main policy goals directs the selection of the most effective interventions in terms of health and economic results~\cite{Hollingsworth2011, Song2022,Pangallo2023,Ash2022,Ohi2020, PADMANABHAN2021102676,NBERw26981,analytical_covid_lockdown_model}. The economic impact of the COVID-19 pandemic varied between rich and poor countries. Although COVID-19 deaths had a slightly larger negative effect on the Gross Domestic Product (GDP) in advanced economies, this difference was not statistically significant. However, lockdown restrictions were found to have a more damaging impact on economic activity in emerging and developing economies~\cite{Gagnon2023,Redlin2022,Liang2021}. It's also suggested that an increase in COVID-19 cases was associated with the introduction of harsher NPIs and lockdown measures could be relaxed once vaccination rates increase~\cite{Redlin2022,Patel2021}.

Many economists have studied the effect of COVID-19 on the economy of nations~\cite{Gagnon2023,GagnonKorea,Deb2020,Eichenbaum2021}. In advanced economies like Korea, where the stringency index was below the median the recession was milder than other advanced economies like the United Kingdom where the stringency was much higher~\cite{GagnonKorea}, they achieved it mostly with very aggressive testing, contact tracing, and enforced quarantines and isolations~\cite{Lim2023,KoreaMinister}. In India, social distancing and containment measures have been effective in reducing the number of COVID-19 cases but have come with economic costs. Social distancing had the most adverse effect on the economy in areas with high urbanization~\cite{Deb2020}.

In this paper we optimize the government policies regarding stringency as it controls both the spread of the disease and the economy. To model the epidemiological data~\cite{WorldometerCorona} we use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the timescale is small enough that it can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we effectively model the disease progression. Our model (SIR with lockdown and time-varying vaccination rate) builds on the foundational SIR model, by accounting for the recovery reached through vaccination~\cite{Marinov2022,MaurciodeCarvalho2023,Thater,Turkyilmazoglu2022,YALADANDA2022101052} and the effects of lockdown~\cite{Hale2021,SIRLockdown,NBERw26981,NBERw26867}. Although the traditional SIR model is a valuable tool for understanding the spread of infectious diseases, it assumes that parameters like the transmission rate ($\beta$) and recovery rate ($\gamma$) are constant over time, which may not always be the case. In this paper, we propose a more sophisticated approach by introducing a time-dependent SIR model~\cite{TimeDependentSIR}, enabling us to account for the changing dynamics of the pandemic due to factors such as lockdowns and vaccination rates. This proposed model effectively addresses the real-world conditions acts as a solution that is both effective and extendable. However, the study has limitations; First, the deterministic SIR model (predecessor to our proposed model) fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}. Secondly, the underreporting of cases during the period selected by our study. Lastly, the reinforcement learning agent should be resistant to how the vaccination rate changes, and different values for $\beta$ and $\gamma$ -- keeping them the same scopes the environment for succumbing to wishful thinking which can be potentially dangerous. Therefore, before an actual deployment of the model, it would be a good measure to introduce stochasticity to these parameters($\beta$ and $\gamma$) and the vaccination rate.

@@

\subsection{Reinforcement Learning}
Given that the government is an agent that takes decisions in a deterministic environment defined above, we use reinforcement learning to model the competing costs of the environment. This environment is known as a Markov decision process (MDP) and is characterized by the Markov property. To possess the Markov property is to create a compact state signal that retains all relevant information from past sensations without requiring the complete history. The Markov property ensures that the probability of transitioning to the next state and receiving a reward depends only on the current state and action, without requiring the entire history~\cite{sutton2018reinforcement}. Our MDP is defined as follows:
% https://arxiv.org/pdf/1708.05866.pdf
\begin{itemize}
    % \item \url{https://jmlr.org/papers/volume22/20-1364/20-1364.pdf} RL Paper that references \url{https://araffin.github.io/slides/rl-tuto-jnrr19/#/13} RL ppt (for our RL notes).
    \item Set of States $\altmathcal{S}$: The state of the environment are described through the descriptors like the normalized GDP ($(GDP_{predicted} - GDP_{min}) / (GDP_{max} - GDP_{min})$), $R_e$, a list of all the previous actions (in changing the stringency) and the proportion of the population that was susceptible ($S$), infected ($I$) and recovered ($R$). The starting states are simply these values at the starting date and no previous actions.
    \item Actions $\altmathcal{A}$: The stringency index variable was analyzed with a sample size of 915. The mean value was approximately $61.96505$, with a standard deviation of $17.66983$. The minimum value was $31.48$, while the maximum value reached $96.3$. And the differences between two consecutive stringencies had a mean of $-0.070919$, and standard deviation of $1.42715$, with the minimum being $-14.36$ and maximum $16.67$. Based on this we define the discrete action space. There are 7 actions for the agent, it can keep the stringency index same, reduce/increase by 2.5, reduce/increase by 5, and reduce/increase by 10 given that the stringency index doesn't exceed 100 or go below 0.
    \item Transition dynamics $\altmathcal{T}\left(\mathbf{s}_{t+1} \mid \mathbf{s}_t, \mathbf{a}_t\right)$ that map a state-action pair at time $t$ onto a distribution of states at time $t+1$. This state transition is defined by the SIR model with lockdown and the model of how stringency index affects the GDP.
    \item Immediate reward $\altmathcal{R}\left(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1}\right)$. The agent observes the state of the environment $\mathbf{s}_t$ at time $t$ and takes an action $\mathbf{a}_t$, after which the state transitions to $\mathbf{s}_{t+1}$ and the agent receives a reward $\mathbf{r}_{t+1}$ as feedback. In \cref{defining_reward_function} we define a reward strategy however it should be noted that this work serves as a framework where the strategy can be easily swapped for another to prioritize different needs.
    \item Discount Factor $\gamma \in[0,1]$, where lower values place more emphasis on immediate rewards. Here, we choose the default discount factor of 0.99.
\end{itemize}

@@

The agent observes the percentage of the population that is susceptible, infected, recovered, and time-varying data like the GDP, and previous actions taken to change the stringency, and $R_e$. Since Stable Baselines3 can support multiple inputs (time-series data, single data points and images) by using Dict Gym space. For data that varies with time (stringency, normalized GDP, $R_e$) we use a simple Long Short Term Memory (LSTM) architecture~\cite{LSTM}. For other data like, the current proportion of the population that is susceptible, infected, and recovered we use a simple fully connected layer. The output from both these networks are concatenated and used by the reinforcement learning agent to train on.

@@

We train the model for $2742$ time steps and some of the best results are presented below.

@@

While non-pharmaceutical interventions (NPIs) can effectively manage the epidemic, they impose economic burdens on developing nations. In \cref{fig:stringency_gdp_developing}, we plot the normalized GDP against the stringency and calculate various metrics like the Pearson correlation coefficient, coefficient of determination ($r^2$), and p-value for three countries (India, Mexico, Brazil) which are Emerging Market and Developing Economies~\cite{IMFCovid} from May 2020 to October 2022. It can be observed from \cref{fig:stringency_gdp_developing} that strict policies have a negative effect on the normalized GDP in developing economies such as India, Mexico, and Brazil. However, this trend is not uniformly seen in advanced economies like the USA, Japan, and Canada as shown in \cref{fig:stringency_gdp_advanced}. In these countries, other factors could be contributing to the decrease in normalized GDP besides the implementation of stricter policies.

@@

In contrast, an alternative output is presented in \cref{fig:175975_parent}, which demonstrates a gradual increase in stringency from October, 2020 till April, 2021. This strategy results in a decline of infections occurring prior to the vaccine's release, and a subsequent cessation of new infections. This approach, however, has implications for the normalized Gross Domestic Product (GDP), as seen by the observed decline in \cref{fig:175975_rl_gdp}. These findings underscore the complexity of managing public health crises and the need for careful strategic planning to balance health outcomes with economic considerations.

>>>>

The need for a nuanced understanding of how interventions impact both health outcomes and economic indicators became increasingly evident, prompting a comprehensive examination by epidemiologists to assist policymakers~\cite{Anderson2020}. The outbreak of COVID-19 has prompted epidemiologists to research on various aspects, including mobility control~\cite{Song2022,Chinazzi2020}, vaccination strategies~\cite{Nguyen2021, Kim2022}, non-pharmaceutical interventions (NPIs) like restricting population movements and gatherings, closing schools and businesses, and requiring masks indoors~\cite{Jalloh2022,Caldwell2021,Ferguson2020}, and financial considerations~\cite{DeFoo2023}. Despite the numerous studies conducted, very few explore how common interventions meet multiple policy objectives or how a precise articulation of the main policy goals directs the selection of the most effective interventions in terms of health and economic results~\cite{Hollingsworth2011, Song2022,Pangallo2023,Ash2022,Ohi2020, PADMANABHAN2021102676,NBERw26981,analytical_covid_lockdown_model}. The economic impact of the COVID-19 pandemic varied between rich and poor countries. Although COVID-19 deaths had a slightly larger negative effect on the Gross Domestic Product (GDP) in advanced economies, this difference was not statistically significant. However, lockdown restrictions were found to have a more damaging impact on economic activity in emerging and developing economies~\cite{Gagnon2023,Redlin2022,Liang2021}. It's also suggested that an increase in COVID-19 cases was associated with the introduction of harsher NPIs and lockdown measures could be relaxed once vaccination rates increase~\cite{Redlin2022,Patel2021}.

Many economists have studied the effect of COVID-19 on the economy of nations~\cite{Gagnon2023,GagnonKorea,Deb2020,Eichenbaum2021}. In advanced economies like Korea, where the stringency index was below the median the recession was milder than other advanced economies like the United Kingdom where the stringency was much higher~\cite{GagnonKorea}, they achieved it mostly with very aggressive testing, contact tracing, and enforced quarantines~\cite{Lim2023,KoreaMinister}. In India, social distancing and containment measures have been effective in reducing the number of COVID-19 cases but have come with economic costs. Social distancing had the most adverse effect on the economy in areas with high urbanization~\cite{Deb2020}.

In this paper we optimize the government policies regarding stringency as it controls both the spread of the disease and the economy. To model the epidemiological data~\cite{WorldometerCorona} we use the simple SIR model without vital dynamics~\cite{Hethcote1989, Hethcote2008, ALLEN2017128} as it is assumed that the timescale is small enough that it can be neglected~\cite{Cooper2020}. By lesioning the model, as opposed to proposing a new mathematical model with more specialized compartments to more accurately represent the actual environment~\cite{Bjrnstad2020, Mwalili2020}, we effectively model the disease progression. Our model (SIR with lockdown and time-varying vaccination rate) builds on the foundational SIR model, by accounting for the recovery reached through vaccination~\cite{Marinov2022,MaurciodeCarvalho2023,Thater,Turkyilmazoglu2022,YALADANDA2022101052} and the effects of lockdown~\cite{Hale2021,SIRLockdown,NBERw26981,NBERw26867}. Although the traditional SIR model is a valuable tool for understanding the spread of infectious diseases, it assumes that parameters like the transmission rate ($\beta$) and recovery rate ($\gamma$) are constant over time, which may not always be the case. In this paper, we propose a more sophisticated approach by introducing a time-dependent SIR model~\cite{TimeDependentSIR}, enabling us to account for the changing dynamics of the pandemic due to factors such as lockdowns and vaccination rates. This proposed model effectively addresses the real-world conditions acts as a solution that is both effective and extendable. However, the study has limitations; First, the deterministic SIR model (predecessor to our proposed model) fails to account for chance in disease spread and lacks confidence intervals on results and while stochastic models incorporate chance, they are typically more challenging to analyze than their deterministic counterparts~\cite{Hethcote2008}. Secondly, the underreporting of cases during the period selected by our study. Lastly, the reinforcement learning agent should be resistant to how the vaccination rate changes, and different values for $\beta$ and $\gamma$ -- keeping them the same scopes the environment for succumbing to wishful thinking which can be potentially dangerous. Therefore, before an actual deployment of the model, it would be a good measure to introduce stochasticity to these parameters ($\beta$ and $\gamma$) and the vaccination rate ($\nu$).

@@

\subsection{Reinforcement Learning}
Given that the government is an agent that takes decisions in a deterministic environment defined above, we use reinforcement learning to model the competing costs of the environment. This environment is known as a Markov decision process (MDP) and is characterized by the Markov property. To possess the Markov property is to create a compact state signal that retains all relevant information from past sensations without requiring the complete history. The Markov property ensures that the probability of transitioning to the next state and receiving a reward depends only on the current state and action, without requiring the entire history~\cite{sutton2018reinforcement}. Our MDP is defined as follows:
% https://arxiv.org/pdf/1708.05866.pdf
\begin{itemize}
    % \item \url{https://jmlr.org/papers/volume22/20-1364/20-1364.pdf} RL Paper that references \url{https://araffin.github.io/slides/rl-tuto-jnrr19/#/13} RL ppt (for our RL notes).
    \item Set of States $\altmathcal{S}$: The state of the environment are described through the descriptors like the normalized GDP, $R_e$, a list of all the previous actions (in changing the stringency) and the proportion of the population that was susceptible ($S$), infected ($I$) and recovered ($R$). The starting states are simply these values at the starting date and no previous actions.
    \item Actions $\altmathcal{A}$: The stringency index variable was analyzed with a sample size of 915. The mean value was approximately $61.96505$, with a standard deviation of $17.66983$. The minimum value was $31.48$, while the maximum value reached $96.3$. And the differences between two consecutive stringencies had a mean of $-0.070919$, and standard deviation of $1.42715$, with the minimum being $-14.36$ and maximum $16.67$. Based on this we define the discrete action space. There are 7 actions for the agent, it can keep the stringency index same, reduce/increase by 2.5, reduce/increase by 5, and reduce/increase by 10 given that the stringency index doesn't exceed 100 or go below 0.
    \item Transition dynamics $\altmathcal{T}\left(\mathbf{s}_{t+1} \mid \mathbf{s}_t, \mathbf{a}_t\right)$ that map a state-action pair at time $t$ onto a distribution of states at time $t+1$. This state transition is defined by the SIR model with lockdown and the model of how stringency index affects the GDP.
    \item Immediate reward $\altmathcal{R}\left(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1}\right)$. The agent observes the state of the environment $\mathbf{s}_t$ at time $t$ and takes an action $\mathbf{a}_t$, after which the state transitions to $\mathbf{s}_{t+1}$ and the agent receives a reward $\mathbf{r}_{t+1}$ as feedback. In \cref{defining_reward_function} we define a reward strategy however it should be noted that this work serves as a framework where the strategy can be easily swapped for another to prioritize different needs.
    \item Discount Factor $\gamma \in[0,1]$, where lower values place more emphasis on immediate rewards. Here, we choose the default discount factor of 0.99.
\end{itemize}

@@

\subsubsection{Deep Reinforcement Learning and Training}\label{drl_and_training}
The agent observes the percentage of the population that is susceptible, infected, recovered, and time-varying data like the GDP, and previous actions taken to change the stringency, and $R_e$. Since Stable Baselines3 can support multiple inputs (time-series data, single data points and images) by using Dict Gym space. For data that varies with time (stringency, normalized GDP, $R_e$) we use a simple Long Short Term Memory (LSTM) architecture~\cite{LSTM}. For other data like, the current proportion of the population that is susceptible, infected, and recovered we use a simple fully connected layer. The output from both these networks are concatenated and used by the reinforcement learning agent to train on. We train the model for $2742$ time steps and some of the best results are presented.

@@

@@

While non-pharmaceutical interventions (NPIs) can effectively manage the epidemic, they impose economic burdens on developing nations. In \cref{fig:stringency_gdp_developing}, we plot the normalized GDP against the stringency and calculate various metrics like the Pearson correlation coefficient, coefficient of determination ($r^2$), and p-value for three countries (India, Mexico, Brazil) which are Emerging Market and Developing Economies~\cite{IMFCovid} from May 2020 to October 2022. It can be observed from \cref{fig:stringency_gdp_developing} that strict policies have a negative effect on the normalized GDP in these economies. However, this trend is not uniformly seen in advanced economies like the USA, Japan, and Canada as shown in \cref{fig:stringency_gdp_advanced}. In these countries, other factors could be contributing to the decrease in normalized GDP besides the implementation of stricter policies.

@@

In contrast, an alternative output is presented in \cref{fig:175975_parent}, which demonstrates a gradual increase in stringency from October, 2020 till April, 2021. This strategy results in a decline of infections occurring prior to the vaccine's release, and a subsequent cessation of new infections. This approach, however, has implications for the normalized GDP, as seen by the observed decline in \cref{fig:175975_rl_gdp}. These findings underscore the complexity of managing public health crises and the need for careful strategic planning to balance health outcomes with economic considerations.
